---
title: "Analysis of Data From Human Activity Recognition"
author: "SC"
date: "Saturday, January 24, 2015"
output: html_document
---
Executive Summary
-----------------
The goal of this project is to use machine learning technique to train a model 
for classification. The data comes from http://groupware.les.inf.puc-rio.br/har. 
Predictor variables come from accelerometers on the belt, forearm, arm, and
dumbell of 6 persons, who were asked to exercise in five different ways (the 
response variable.) The data has 19622 entries with 159 predictors and one
response.   

Pre-processing
---------------

```{r, results='hide'}
    #load packages and set random seed (assuming caret and e1071 installed)
    library(caret)
    library(e1071)
    set.seed(1234)    
    #load data set
    train <- read.csv("pml-training.csv")
    # take a look at data
    summary(train)
    # replace empty cells with NA
    train1 <- apply(train, 2, function(x) ifelse(x =="", NA, x))
    # sum number of NAs in each column
    train2 <- apply(train1, 2, function(x)sum(is.na(x)))
    #extract names of columns with NAs
    names_to_remove<- names(train2[which(train2>0)])
    #remove columns with NAs
    train3 <- train1[, !(colnames(train1) %in% names_to_remove)]
    # convert matrix train3 to data.frame train4
    train4 <- as.data.frame(train3)
    # extract names of other columns to be removed
    names_to_remove2 <- c("X","user_name","raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp","new_window")
    # remove those columns
    train6 <- train4[, !(colnames(train4) %in% names_to_remove2)]
    # put the response variable in y
    y <- train6$classe
    # delete classe column
    train6$classe <-NULL
    # convert the data set into numeric
    train6 <- apply(train6, 2, as.numeric)
    #combine y with train6
    train7 <- as.data.frame(cbind(train6,y))
    # change column "y" to "classe"
    names(train7)[54] <- "classe"
    # convert column classe to factor
    train7$classe <- as.factor(train7$classe)
    # divide data into training and testing sets
    inTrain <- createDataPartition(y=train$classe, p=0.75, list=FALSE)
    training <- train7[inTrain,]
    testing <- train7[-inTrain,] 
```
After downloading the data, I look at the summary of the data. I notice a large
number of columns are empty. So I first fill the empty cells with NA, extract 
column names with NAs and remove them from the data. I also take out the first 
six columns that seem to have irrelevant information. Then I partition the data 
so that 3/4 is for training and validation, and the rest testing. During the
pre-processing step, for a couple of times, I have to convert the data from 
matrix, which is a result of selecting a subset of columns from a data.frame, 
back to data.frame. At the end, I also have to convert the response variable to
factor. The factor levels become 1-5 instead of A-E during preprocessing. After 
pre-processing, there are 53 predictors and one response.

Model Selection
---------------
I run the training set with Random Forest because it is one of the top performing
algorithms for classification. I use 10-fold cross-validation option. Given the 
accuracy of the model selected in CV, the out-of sample error should be closed
to 0.2%. The selected model is then used to make prediction on
the testing set.

```{r}
    #set control parameters 
    control <- trainControl(method = "cv", number=10, savePred=TRUE)
    #training model
    model_rf <- train(classe ~., data=training, method="rf",trControl = control)
    #predict
    print(model_rf)
    pred <-predict(model_rf, newdata=testing)
    #check accuracy
    confusionMatrix(pred, testing$classe)$overall
```

Prediction
----------
The selected model is used to predict 20 outcomes given in the project.
```{r}
    test <- read.csv("pml-testing.csv")
    test1 <- apply(test, 2, function(x) ifelse(x =="", NA, x))
    test3 <- test1[, !(colnames(test1) %in% names_to_remove)]
    test4 <- as.data.frame(test3)
    names_to_remove2 <- c("X","raw_timestamp_part_1", "raw_timestamp_part_2", "cvtd_timestamp")
    test5 <- test4[, !(colnames(test4) %in% names_to_remove2)]
    test6 <- test5[, 3:dim(test5)[2]]
    test6$problem_id <-NULL
    # convert the data set into numeric
    test6 <- apply(test6, 2, as.numeric)
    pred <-predict(model_rf, newdata=test6)   
    print(pred)
```
Conclusion
----------
The prediction accuracy of the test set is close to 100%. All twenty outcomes 
are predicted correctly. 
